---
title: "India's Gross Domestic Product and Its Components"
author: "Udayan Roy"
date: "`r Sys.Date()`"
format: html
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE

library(tidyverse)
library(readxl)
library(lubridate)

quarterly <- read_xlsx(
  path = "india-gdp.xlsx",
  sheet = "Components_Q_2011_12_prices",
  col_names = TRUE
) |>
  mutate(Date = as.Date(paste0(as.numeric(substr(Year, start = 1, stop = 4)), "-", case_when(Quarter == 1 ~ "01",
                           Quarter == 2 ~ "04",
                           Quarter == 3 ~ "07",
                           Quarter == 4 ~ "10"), "-01"))) |>
  select(c(Year, Quarter, GDP:Discrepancies,Date))

share <- function(x) {
  100*x/quarterly$GDP
}

share_incr <- function(x, y){
  100*(x - lag(x, 4))/(y - lag(y, 4))
}

growth <- function(x){
  100*(x - lag(x, 4))/lag(x, 4)
}



quarterly_gdp <- quarterly |>
  mutate(I = GFCF + CIS + Valuables,
         I_share = share(I),
         C_share = share(PFCE),
         G_share = share(GFCE),
         NX = Exports - Imports,
         NX_share = share(NX),
         EX_share = share(Exports),
         IM_share = share(Imports),
         D_share = share(Discrepancies),
         Total = I_share + C_share + G_share + NX_share + D_share,
         C_share_incr = share_incr(PFCE, GDP),
         I_share_incr = share_incr(I, GDP),
         G_share_incr = share_incr(GFCE, GDP),
         NX_share_incr = share_incr(NX, GDP),
         EX_share_incr = share_incr(Exports, GDP),
         IM_share_incr = share_incr(Imports, GDP),
         D_share_incr = share_incr(Discrepancies, GDP),
         Total_share_incr = C_share_incr + I_share_incr + G_share_incr + NX_share_incr + D_share_incr,
         C_growth = growth(PFCE),
         G_growth = growth(GFCE),
         EX_growth = growth(Exports),
         IM_growth = growth(Imports),
         I_growth = growth(I),
         GDP_growth = growth(GDP),
         NX_growth = sign(NX - lag(NX, 4))*100*
(((abs(lag(NX, 4)) + abs(NX - lag(NX, 4)))/abs(lag(NX, 4)))-1)
) 

## Average annual growth over data period

date.begin <- quarterly_gdp$Date[year(quarterly_gdp$Date) == year(min(quarterly_gdp$Date)) & quarter(quarterly_gdp$Date) == quarter(max(quarterly_gdp$Date))]
date.end <- max(quarterly_gdp$Date)

growth_mean <- function(x){
  (((x[quarterly_gdp$Date == date.end]/x[quarterly_gdp$Date == date.begin])^(1/(year(date.end) - year(date.begin)))) - 1)*100
}

GDP_growth_mean <- growth_mean(quarterly_gdp$GDP)
C_growth_mean <- growth_mean(quarterly_gdp$PFCE)
I_growth_mean <- growth_mean(quarterly_gdp$I)
G_growth_mean <- growth_mean(quarterly_gdp$GFCE)
EX_growth_mean <- growth_mean(quarterly_gdp$Exports)
IM_growth_mean <- growth_mean(quarterly_gdp$Imports)
NX_growth_mean <-  sign((quarterly_gdp$NX[quarterly_gdp$Date == date.end] - quarterly_gdp$NX[quarterly_gdp$Date == date.begin]))*
  (((((abs(quarterly_gdp$NX[quarterly_gdp$Date == date.begin]) +
  abs(quarterly_gdp$NX[quarterly_gdp$Date == date.end] - quarterly_gdp$NX[quarterly_gdp$Date == date.begin]))/
  abs(quarterly_gdp$NX[quarterly_gdp$Date == date.begin]))^(1/(year(date.end) - year(date.begin)))) - 1)*100)


```

## The Latest Quarter At A Glance

Data on India's Gross Domestic Product for Quarter `r max(quarterly_gdp$Quarter[quarterly_gdp$Year==max(quarterly_gdp$Year)])` of the `r max(quarterly_gdp$Year)` fiscal year has just been published. Here are some highlights:

-   Inflation-adjusted Gross Domestic Product was INR `r format(quarterly_gdp$GDP[quarterly_gdp$Date == max(quarterly_gdp$Date)], scientific = FALSE, nsmall = 2, big.mark=",")` crores at constant 2011-12 prices. (A crore is ten million or 10<sup>7</sup>.)\
-   This represents a growth of [`r round(quarterly_gdp$GDP_growth[quarterly_gdp$Date==max(quarterly_gdp$Date)],2)` percent]{style="color: red;"} compared to the same quarter in the previous fiscal year. (The available quarterly data spans Quarter `r min(quarterly_gdp$Quarter[quarterly_gdp$Date==min(quarterly_gdp$Date)])` of the `r min(quarterly_gdp$Year)` fiscal year through Quarter `r max(quarterly_gdp$Quarter[quarterly_gdp$Date==max(quarterly_gdp$Date)])` of the `r max(quarterly_gdp$Year)` fiscal year. That's 13 years. Over the past 13 years, GDP has grown [`r round(GDP_growth_mean,2)` percent]{style="color: blue;"} annually.)
-   Consumption spending grew [`r round(quarterly_gdp$C_growth[quarterly_gdp$Date==max(quarterly_gdp$Date)],2)` percent]{style="color: red;"}. (Over the past 13 years, consumption spending has grown [`r round(C_growth_mean,2)` percent]{style="color: blue;"} annually.)
-   Investment spending grew [`r round(quarterly_gdp$I_growth[quarterly_gdp$Date==max(quarterly_gdp$Date)],2)` percent]{style="color: red;"}. (Over the past 13 years, investment spending has grown [`r round(I_growth_mean,2)` percent]{style="color: blue;"} annually.)
-   Government purchases grew [`r round(quarterly_gdp$G_growth[quarterly_gdp$Date==max(quarterly_gdp$Date)],2)` percent]{style="color: red;"}. (Over the past 13 years, government purchases has grown [`r round(G_growth_mean,2)` percent]{style="color: blue;"} annually.)\
-   Exports grew [`r round(quarterly_gdp$EX_growth[quarterly_gdp$Date==max(quarterly_gdp$Date)],2)` percent]{style="color: red;"}. (Over the past 13 years, exports has grown [`r round(EX_growth_mean,2)` percent]{style="color: blue;"} annually.)
-   Imports grew [`r round(quarterly_gdp$IM_growth[quarterly_gdp$Date==max(quarterly_gdp$Date)],2)` percent]{style="color: red;"}. (Over the past 13 years, imports has grown [`r round(IM_growth_mean,2)` percent]{style="color: blue;"} annually.)
-   Net Exports grew [`r round(quarterly_gdp$NX_growth[quarterly_gdp$Date==max(quarterly_gdp$Date)],2)` percent]{style="color: red;"}. (Over the past 13 years, net exports has grown [`r round(NX_growth_mean,2)` percent]{style="color: blue;"} annually.)
-   To sum up the latest quarter, **GDP growth was above the long-run average. Above-average growth in both consumption and investment spending was the highlight. Government purchases, which has been weak for several past quarters, actually shrank. Exports and imports both grew faster than average, with exports growth outpacing growth in imports. Nothing stuck out.**

## Gross Domestic Product, Level

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE

options(scipen = 999)
library(scales)

graphit <- function(var, title, subtitle){
  ggplot(data = quarterly_gdp, mapping = aes(x = Date)) +
  geom_line(mapping = aes(y = var)) +
  geom_point(mapping = aes(y = var)) +
  labs (x = NULL, y = NULL, 
        title = title,
        subtitle = subtitle,
        caption = "Each yearly tick mark represents April 1 (which is the first day of the April 1 -- March 31 fiscal year).\nThe data are from the Government of India.") +
  theme_bw() + 
  scale_y_continuous(labels = scales::comma) +
  scale_x_date(labels = date_format("%Y"), date_breaks = "2 years")
}

graphit(var = quarterly_gdp$GDP, title = "India's GDP", subtitle = "Quarterly GDP is in tens of millions (crores) of constant 2011-12 Rupees.")

```

*Comment*: Just eyeball the pre-pandemic and post-pandemic parts of the chart for a second or two. Note the increased volatility *after* the pandemic, with sharp declines in the first quarter of each year followed by equally sharp increases in the subsequent quarters. I wonder why this is so. Things seem to be returning to the pre-pandemic pattern (of mild increases). However, the question remains: *Why has the first quarter (April -- June) been so far below the pre-pandemic trend after the pandemic? More puzzlingly, why has fourth quarter (January -- March) been so close to the pre-pandemic trend?*

## Gross Domestic Product, Growth Rate

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE

subtitle_growth <- function(index_name, index, index_mean){
	paste("In the latest quarter,", index_name, "grew", round(index[quarterly_gdp$Date==max(quarterly_gdp$Date)],2), "percent. \nThe dashed line shows the average since Quarter", min(quarterly_gdp$Quarter[quarterly_gdp$Date==min(quarterly_gdp$Date)]), "of the",  min(quarterly_gdp$Year), "fiscal year:", round(index_mean,2), "percent.")
}

subtitle_share <- function(index_name, index, index_mean){
	paste("In the latest quarter,", index_name, "was", round(index[quarterly_gdp$Date==max(quarterly_gdp$Date)],2), "percent of GDP. \nThe dashed line shows the average since Quarter", min(quarterly_gdp$Quarter[quarterly_gdp$Date==min(quarterly_gdp$Date)]), "of the",  min(quarterly_gdp$Year), "fiscal year:", round(index_mean,2), "percent.")
}

subtitle_share_incr <- function(index_name, index, index_mean){
	paste("In the latest quarter, the increase in", index_name, "was\n", round(index[quarterly_gdp$Date==max(quarterly_gdp$Date)],2), "percent of the increase in GDP. \nThe dashed line shows the average since Quarter", min(quarterly_gdp$Quarter[quarterly_gdp$Date==min(quarterly_gdp$Date)]), "of the",  min(quarterly_gdp$Year), "fiscal year:", round(index_mean,2), "percent.")
}


graphit2 <- function(index, title, subtitle, index_mean){
  ggplot(data = quarterly_gdp, mapping = aes(x = Date)) +
  geom_line(mapping = aes(y = index)) +
  geom_point(mapping = aes(y = index))  +
  geom_hline(aes(yintercept = index_mean), linetype = "dashed") +
  labs (x = NULL, y = NULL, 
        title = title,
        subtitle = subtitle,
        caption = "Each yearly tick mark represents April 1 (which is the first day of the April 1 -- March 31 fiscal year).\nThe data are from the Government of India.") +
  theme_bw() + 
  scale_y_continuous(labels = scales::comma) +
  scale_x_date(labels = date_format("%Y"), date_breaks = "2 years")
}

mysubtitle <- subtitle_growth("(inflation-adjusted) GDP", quarterly_gdp$GDP_growth, GDP_growth_mean)

graphit2(index = quarterly_gdp$GDP_growth, index_mean = GDP_growth_mean, title = "Growth Rate of India's GDP", subtitle = mysubtitle)

```

## The Components of GDP (*C*, *I*, *G*, and *NX*)

GDP is the rupee value of all production, roughly speaking. Consequently, GDP is also total income and total spending. Total spending is usually broken down into four *components*:

-   consumption spending (*C*, which is present oriented),
-   investment spending (*I*, which is future oriented),
-   government purchases (*G*), and
-   net exports (*NX* = *Exports* - *Imports*).

Therefore, *GDP* = *C* + *I* + *G* + *NX*. It is important to know which *kind* of spending -- that is, which component of GDP -- contributes how much to GDP and its growth. Policy makers may be able to design better macroeconomic policies if they can spot the component that is causing trouble.

There are a dozen charts below, three for each of the four components.

-   The *first* of the three charts for a particular component shows that component's *growth rate* over the previous year.
-   The *second* shows that component's share of total spending (that is, its *share of GDP*) during each quarter.
-   And the *third* shows the increase in that component over the previous year as a share of the increase in total spending over the previous year (that is, *its increase as a share of the increase in GDP*).

Each chart shows the relevant values for each quarter as well as the relevant long-term average from 2011 onwards (shown with a dashed horizontal line).

### Consumption Spending

*Importance*: Consumption spending was `r round(quarterly_gdp$C_share[quarterly_gdp$Date==max(quarterly_gdp$Date)],2)` percent of GDP in the latest quarter (Quarter `r max(quarterly_gdp$Quarter[quarterly_gdp$Year==max(quarterly_gdp$Year)])` of the `r max(quarterly_gdp$Year)` fiscal year). This can be compared with Consumption's `r round(mean(quarterly_gdp$C_share, na.rm = TRUE),2)` percent share of GDP over the full data period (spanning Quarter `r min(quarterly_gdp$Quarter[quarterly_gdp$Date==min(quarterly_gdp$Date)])` of the `r min(quarterly_gdp$Year)` fiscal year through Quarter `r max(quarterly_gdp$Quarter[quarterly_gdp$Date==max(quarterly_gdp$Date)])` of the `r max(quarterly_gdp$Year)` fiscal year).

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE

mysubtitle <- subtitle_growth("(inflation-adjusted) consumption spending", quarterly_gdp$C_growth, C_growth_mean)

graphit2(index = quarterly_gdp$C_growth, index_mean = C_growth_mean, title = "Growth Rate of Consumption Spending", subtitle = mysubtitle)

mysubtitle <- subtitle_share("(inflation-adjusted) consumption spending", quarterly_gdp$C_share, mean(quarterly_gdp$C_share, na.rm = TRUE))

graphit2(index = quarterly_gdp$C_share, index_mean = mean(quarterly_gdp$C_share, na.rm = TRUE), title = "Share of Consumption Spending in GDP", subtitle = mysubtitle)

mysubtitle <- subtitle_share_incr("(inflation-adjusted) consumption spending", quarterly_gdp$C_share_incr, mean(quarterly_gdp$C_share_incr, na.rm = TRUE))

graphit2(index = quarterly_gdp$C_share_incr, index_mean = mean(quarterly_gdp$C_share_incr, na.rm = TRUE), title = "Increase in Consumption Spending as a share of increase in GDP", subtitle = mysubtitle)


```

### Investment Spending

*Importance*: Investment spending was `r round(quarterly_gdp$I_share[quarterly_gdp$Date==max(quarterly_gdp$Date)],2)` percent of GDP in the latest quarter. This can be compared with Investment's `r round(mean(quarterly_gdp$I_share, na.rm = TRUE),2)` percent share of GDP over the full data period.

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE

mysubtitle <- subtitle_growth("(inflation-adjusted) investment spending", quarterly_gdp$I_growth, I_growth_mean)

graphit2(index = quarterly_gdp$I_growth, index_mean = I_growth_mean, title = "Growth Rate of Investment Spending", subtitle = mysubtitle)

mysubtitle <- subtitle_share("(inflation-adjusted) investment spending", quarterly_gdp$I_share, mean(quarterly_gdp$I_share, na.rm = TRUE))

graphit2(index = quarterly_gdp$I_share, index_mean = mean(quarterly_gdp$I_share, na.rm = TRUE), title = "Share of Investment Spending in GDP", subtitle = mysubtitle)

mysubtitle <- subtitle_share_incr("(inflation-adjusted) investment spending", quarterly_gdp$I_share_incr, mean(quarterly_gdp$I_share_incr, na.rm = TRUE))

graphit2(index = quarterly_gdp$I_share_incr, index_mean = mean(quarterly_gdp$I_share_incr, na.rm = TRUE), title = "Increase in Investment Spending as a share of increase in GDP", subtitle = mysubtitle)

```

### Government Purchases

*Importance*: Government Purchases was `r round(quarterly_gdp$G_share[quarterly_gdp$Date==max(quarterly_gdp$Date)],2)` percent of GDP in the latest quarter. This can be compared with Government Purchases's `r round(mean(quarterly_gdp$G_share, na.rm = TRUE),2)` percent share of GDP over the full data period.

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE

mysubtitle <- subtitle_growth("(inflation-adjusted) government spending", quarterly_gdp$G_growth, G_growth_mean)

graphit2(index = quarterly_gdp$G_growth, index_mean = G_growth_mean, title = "Growth Rate of Government Spending", subtitle = mysubtitle)

mysubtitle <- subtitle_share("(inflation-adjusted) government spending", quarterly_gdp$G_share, mean(quarterly_gdp$G_share, na.rm = TRUE))

graphit2(index = quarterly_gdp$G_share, index_mean = mean(quarterly_gdp$G_share, na.rm = TRUE), title = "Share of Government Spending in GDP", subtitle = mysubtitle)

mysubtitle <- subtitle_share_incr("(inflation-adjusted) investment spending", quarterly_gdp$G_share_incr, mean(quarterly_gdp$G_share_incr, na.rm = TRUE))

graphit2(index = quarterly_gdp$G_share_incr, index_mean = mean(quarterly_gdp$G_share_incr, na.rm = TRUE), title = "Increase in Government Spending as a share of increase in GDP", subtitle = mysubtitle)

```

### Net Exports

*Importance*: Net Exports was `r round(quarterly_gdp$NX_share[quarterly_gdp$Date==max(quarterly_gdp$Date)],2)` percent of GDP in the latest quarter. This can be compared with Net Exports's `r round(mean(quarterly_gdp$NX_share, na.rm = TRUE),2)` percent share of GDP over the full data period.

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE

mysubtitle <- subtitle_growth("(inflation-adjusted) net exports", quarterly_gdp$NX_growth, NX_growth_mean)

graphit2(index = quarterly_gdp$NX_growth, index_mean = NX_growth_mean, title = "Growth Rate of Net Exports", subtitle = mysubtitle)

mysubtitle <- subtitle_share("(inflation-adjusted) net exports", quarterly_gdp$NX_share, mean(quarterly_gdp$NX_share, na.rm = TRUE))

graphit2(index = quarterly_gdp$NX_share, index_mean = mean(quarterly_gdp$NX_share, na.rm = TRUE), title = "Share of net exports in GDP", subtitle = mysubtitle)

mysubtitle <- subtitle_share_incr("(inflation-adjusted) net exports", quarterly_gdp$NX_share_incr, mean(quarterly_gdp$NX_share_incr, na.rm = TRUE))

graphit2(index = quarterly_gdp$NX_share_incr, index_mean = mean(quarterly_gdp$NX_share_incr, na.rm = TRUE), title = "Increase in Net Exports as a share of increase in GDP", subtitle = mysubtitle)


```

#### Exports

*Importance*: Exports was `r round(quarterly_gdp$EX_share[quarterly_gdp$Date==max(quarterly_gdp$Date)],2)` percent of GDP in the latest quarter. This can be compared with Exports's `r round(mean(quarterly_gdp$EX_share, na.rm = TRUE),2)` percent share of GDP over the full data period.

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE

mysubtitle <- subtitle_growth("(inflation-adjusted) exports", quarterly_gdp$EX_growth, EX_growth_mean)

graphit2(index = quarterly_gdp$EX_growth, index_mean = EX_growth_mean, title = "Growth Rate of Exports", subtitle = mysubtitle)

mysubtitle <- subtitle_share("(inflation-adjusted) exports", quarterly_gdp$EX_share, mean(quarterly_gdp$EX_share, na.rm = TRUE))

graphit2(index = quarterly_gdp$EX_share, index_mean = mean(quarterly_gdp$EX_share, na.rm = TRUE), title = "Share of exports in GDP", subtitle = mysubtitle)

mysubtitle <- subtitle_share_incr("(inflation-adjusted) exports", quarterly_gdp$EX_share_incr, mean(quarterly_gdp$EX_share_incr, na.rm = TRUE))

graphit2(index = quarterly_gdp$EX_share_incr, index_mean = mean(quarterly_gdp$EX_share_incr, na.rm = TRUE), title = "Increase in Exports as a share of increase in GDP", subtitle = mysubtitle)


```

#### Imports

*Importance*: Imports was `r round(quarterly_gdp$IM_share[quarterly_gdp$Date==max(quarterly_gdp$Date)],2)` percent of GDP in the latest quarter. This can be compared with Imports's `r round(mean(quarterly_gdp$IM_share, na.rm = TRUE),2)` percent share of GDP over the full data period.

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE

mysubtitle <- subtitle_growth("(inflation-adjusted) imports", quarterly_gdp$IM_growth, IM_growth_mean)

graphit2(index = quarterly_gdp$IM_growth, index_mean = IM_growth_mean, title = "Growth Rate of Imports", subtitle = mysubtitle)

mysubtitle <- subtitle_share("(inflation-adjusted) imports", quarterly_gdp$IM_share, mean(quarterly_gdp$IM_share, na.rm = TRUE))

graphit2(index = quarterly_gdp$IM_share, index_mean = mean(quarterly_gdp$IM_share, na.rm = TRUE), title = "Share of imports in GDP", subtitle = mysubtitle)

mysubtitle <- subtitle_share_incr("(inflation-adjusted) imports", quarterly_gdp$IM_share_incr, mean(quarterly_gdp$IM_share_incr, na.rm = TRUE))

graphit2(index = quarterly_gdp$IM_share_incr, index_mean = mean(quarterly_gdp$IM_share_incr, na.rm = TRUE), title = "Increase in Imports as a share of increase in GDP", subtitle = mysubtitle)

```
